Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
updated alternative methods

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide statistics to measure the strength of extreme heterosis of each gene.

{\tt edgeR} is designed to apply a negative binomial loglinear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a pseudo-probability of heterosis for each gene. We first computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then we used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2. For each gene, if the estimated mean expression level of the hybrid is not between the estimated mean expression levels of the parents, we set the pseudo-probability of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid. Otherwise, the pseudo-probability is zero.

The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}. In a technique similar to our application of {\tt edgeR}, we used used the posterior probabilities of these hypotheses to construct a pseudo-probability of heterosis for each gene. We set this pseudo-probability to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated with {\tt edgeR} is between those of the parents. Otherwise, the pseudo-probability is the sum of the posterior probabilities of $H_2$ and $H_5$. Note that $H_2$ and $H_5$ are the only hypotheses that allow for heterosis.

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$. Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

>>>>

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.

{\tt edgeR} is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2. For each gene, if the estimated mean expression level of the hybrid is between the estimated mean expression levels of the parents, we set the measure of heterosis to zero. Otherwise, we set the measure of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.

The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct. In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated using {\tt edgeR} is between those of the parents. Otherwise, the measure is the sum of the posterior probabilities of $H_2$ and $H_5$, the only hypotheses that allow for extreme heterosis.

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a measure of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities. Specifically, the measure of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
